---
title: "OuriGAMi"
output: github_document 
---

This Rmd runs OuriGAMi, a tool which uses a generalized additive model to identify genes that are correlated with dormancy pseudotime and cell cycle pseudotime. OuriGAMi requires path to adata, ouroborous pseudotime result, and metadata file (file with cell_id to condition). 

```{r setup}
adata_path <- '/projects/steiflab/scratch/glchang/livBC/downsample/rosano.h5ad'
obo_path <- '/projects/steiflab/scratch/glchang/data/Rosano_BC/ouroboros_embeddings_pseudotimes.csv'

# Can be NA if no experimental condition to contrast
meta_path <- '/projects/steiflab/scratch/glchang/livBC/downsample/rosano_meta.csv'
```

### Installation
OuriGAMi requires the following R packages: 
```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) 
  install.packages("BiocManager") 

install.packages(c("mgcv","furrr","future","dplyr","readr","progressr","ggplot2","purrr","reticulate", "patchwork"))
BiocManager::install(c("scran","scuttle","zellkonverter"))
```


#### Set up enrivoment and import data
This pipeline depends on:

1. The R helper script: OuriGAMi_utils.R (use absolute path if util file is not found)
2. ouroboros_env python path (used via reticulate)

To find the correct Python path, run in terminal and copy output to PYTHOHN_ENV:
readlink -f $(which python)
```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
PYTHON_ENV <- "/home/glchang/miniconda3/envs/ouroboros_env/bin/python3.6"

Sys.setenv(RETICULATE_PYTHON = PYTHON_ENV)

# Run .rs.restartR() if you get an error below and rerun whole script
suppressPackageStartupMessages({
  library(mgcv)
  library(zellkonverter)
  library(furrr)
  library(future)
  library(dplyr)
  library(readr)
  library(progressr)
  library(ggplot2)
  library(purrr)
  library(reticulate)
  library(scuttle)
  library(scran)
  library(patchwork)
})

# If this fails, replace with full path to the file
source("OuriGAMi_utils.R")
```

This will read in the adata and convert it to SingleCellExperiment object. The metadata will also be imported, make sure to select the correct column name for your cell id and experimental group.
```{r}
# Can alternative read in file using matrix.mtx, obs.csv, var.csv and build SingleCellExperiment object manually
sce <- read_adata(adata_path)

# Expect 2-3 columns: "cell" (cell_id), "exp" (experimental condition) and optionally "patient_id" to use it as co-variate
# Can also be retrieved using colData(sce) if stored in SingleCellExperiment
# Set meta as NA if you aren't comparing between conditions
meta <- read_delim(meta_path, show_col_types = FALSE) %>%
  `colnames<-`(c("cell", 'exp', 'patient_id'))

head(meta)
```

Next filter sce to remove genes with low expression. Optionally, sce can be further filtered to run OurGAMi on HVG only and/or specific genes of interest. Afterwards, read in metadata and ouroboros data. 

```{r, include=TRUE} 
assay_type = 'X' # Layer for raw counts in SingleCellExperiment object
min_prop_cell= 0.05 # Filter for proportion of cells expressing gene

n_hvg = 250 # Number of HVG, NA to not filter by HVG
genes = NA

# Filter low expression genes
if (!all(is.na(min_prop_cell)) & min_prop_cell != 0){
  keep <- rowSums(assay(sce, assay_type) > 0) / ncol(sce) >= min_prop_cell
  sce <- sce[keep, ]
}

# Filter for genes of interest
if (all(is.na(genes))) {
    genes <- rownames(sce)
} else {
    genes <- intersect(genes, rownames(sce))
}

sce <- sce[, colSums(assay(sce, assay_type)) > 0]
sce <- logNormCounts(sce, assay.type=assay_type)

# Filter for HVG
if (!all(is.na(n_hvg)) & n_hvg != 0){
  gene_var <- modelGeneVar(sce, assay.type = "logcounts")
  hvgs <- getTopHVGs(gene_var, n = n_hvg, var.threshold = -Inf)
  genes <- intersect(hvgs, genes)
}

files <- read_files(sce, meta, obo_path, genes)
```

## Get gene score (optional)
Rather than running GAM model gene expression vs pseudotime, we can run gene set score vs pseudotime. This will require named list where name is gene set name and vector is gene name. Set HVG = 0 and n_hvg = 0
```{r eval=FALSE, include=TRUE}
signature <- read_delim('hallmark_genes.csv', show_col_types = FALSE)
gene_sets <- split(signature$gene_name, signature$pathway)

gene_set_mat <- get_gene_set(files, gene_sets)

files$expr_mat <- gene_set_mat$score_mat
files$genes <- gene_set_mat$gene_sets
```

## GAM paramters
GAM parameters can be specified below. The main parameter is k which controls the smoothness or waviness of the model. The family function should be set to gaussian() for log normalized or gene scores and nb() (negative binomial) should be used for raw counts. 
```{r}
threads=16
BIN_SIZE <- 30

reference_exp <- 'start'  # Reference condition (Must be one of the condition in meta; NA if meta doesn't exist)
k=5                                  # GAM model parameter that controls smoothness/waviness of model
family_function <- gaussian()        # use nb() if raw counts
```

## Dormancy Depth
This runs GAM model for every genes in files$genes vs dormancy pseudotime. This will take a while to run, decrease number of genes or increase threads to speed up.
```{r}
type <- 'dormancy'
binned_files_dorm <- bin_files(files, type, BIN_SIZE)
dormancy_metric <- run_GAM(binned_files_dorm, type, files$genes, threads=threads, k=k, family_function=family_function, ref_exp=reference_exp, cyclic_cubic=F)
dormancy_metric 
```


## Cell Cycle Pseudotime
This runs GAM model for every genes in files$genes vs cell cycle and dormancy pseudotime. cyclic_cubic can be set to TRUE to have model wrap around (G1 starts where G2M ends).
```{r}
type <- 'cell_cycle'
binned_files_cc <- bin_files(files, type, BIN_SIZE)
cc_metric <- run_GAM(binned_files_cc, type, files$genes, threads=threads, k=k, family_function=family_function, ref_exp=reference_exp, cyclic_cubic=F)
cc_metric
```


## Both dormancy and cell cycle pseudotime combined
This runs GAM model for every genes in files$genes vs pseudotime (combined cell cycle and dormancy). 
```{r}
type <- 'all'
binned_files_all <- bin_files(files, type, BIN_SIZE)
all_metric <- run_GAM(binned_files_all, type, files$genes, threads=threads, k=k, family_function=family_function, ref_exp=reference_exp, cyclic_cubic=F)
all_metric
```



## GAM output
Both dormancy_metric and cc_metric outputs statistics of GAM models. 
Model Fit Quality

- **R2:** Variance in gene counts explained by the model
- **deviance:** Proportion of deviance explained by the model

Pseudotime Main Effect (All cells aggregated)

- **main_p:** Tests whether gene expression changes significantly along pseudotime for all cells
- **main_edf:** Complexity of curve for of all cells 

Per-Experiment Metrics

- **`{exp}`_p:** Significance of `{exp}` curve shape differing from all other cells
- **`{exp}`_edf:** Complexity of `{exp}` curve
- **`{exp}`_contrast:** Different baseline level compared to reference_exp
- **`{exp}`_peak:** Region of pseudotime where `{exp}` curve peaked
 
Patient Covariate (optional)

- **patient_p:** Tests whether expression differs significantly across patients (vertical shift)
- **patient_edf:** Magnitude of expression variability across patients (0 = patients don't differ, higher = greater between-patient differences)

## Plot GAM for individual gene
### `plot_gene()`

Plots gene expression against pseudotime, with a fitted GAM trend line and
phase-region annotations along the top of the plot.

**Parameters**

- **`gene`** — Name of the gene to plot. 

- **`binned_files`** — Generated above using bin_files. Named `binned_files_dorm`, `binned_files_cc`, `binned_files_all` in this file

- **`colour_by_patient`** — Colour points by patient_id (default FALSE); only applicable if patient_id is provided in the meta file."

- **`palette`** — Named character vector of hex colors, or `NA` (default palette). If colour_by_patient = TRUE, must include names for both exp and patient_id levels.

**Returns:** a `ggplot` object.

```{r}
gene_of_interest <- 'MT2A'
plot_gene(gene_of_interest, binned_files_dorm)
```


Can colour points based on patient_id if provided in meta.

```{r}
gene_of_interest <- 'IGFBP5'
plot_gene(gene_of_interest, binned_files_cc, colour_by_patient = T)
```

Colour palette can be provided to plot_gene function. 

```{r}
gene_of_interest <- 'TPD52L1'
plot_gene(gene_of_interest, binned_files_all, palette = c('Awakening'='#7C606B', 'month_1'='#C46BAE', 'month_2'='#EB7BC0', 'start'='#EDA4BD' ))
```


## Plot top genes 

## `plot_multi_genes()`
Plots expression against pseudotime for multiple genes at once, faceted one
panel per gene, each with a fitted GAM trend line and phase-region annotations
along the top.

**Parameters**

- **`genes_lst`** — Character vector of gene names to plot, one facet per gene.
- **`binned_files`** — Generated above using `bin_files`. Named `binned_files_dorm`, `binned_files_cc`, `binned_files_all` in this file.
- **`type`** — Type of analysis (`cell_cycle`, `dormancy` or `all`)
- **`legend`** — Logical, default `TRUE`. Whether to show the color legend.
- **`palette`** — Named character vector of hex colors, or `NULL` (default palette).

**Returns:** a `ggplot` object, faceted by gene with free y-axis scales.

The cc_metric and dormancy_metric can be filtered and arrange to find genes of interest. The genes can be filtered based on p-value, edf, peak region. The top genes can be plotted using `plot_multi_genes` function.

```{r fig.width=10, fig.height=7}
top_genes <- cc_metric %>%
  dplyr::filter(main_p < 0.05) %>%
  dplyr::filter(main_peak_region == 'G1') %>%
  arrange(desc(R2)) %>%
  dplyr::slice(1:12) %>%
  pull(gene)

plot_multi_genes(top_genes, binned_files_cc, 'cell_cycle', legend=T)
```

You can filter based on specific condition (e.g Awakening_p < 0.05), peak regions of each condition (e.g Awakening_peak_region == 'deep_dorm') to find genes of interest in specific condition and region. 
```{r fig.width=10, fig.height=7}
top_genes <- dormancy_metric %>%
  dplyr::filter(main_p < 0.05) %>%
  dplyr::filter(Awakening_p < 0.05) %>%
  dplyr::filter(Awakening_peak_region == 'deep_dorm') %>%
  arrange(desc(R2)) %>%
  dplyr::slice(1:12) %>%
  pull(gene)

plot_multi_genes(top_genes, binned_files_dorm, 'dormancy', legend=T, palette = c('Awakening'='#111184', 'month_1'='#C46BAE', 'month_2'='#EB7BC0', 'start'='#EDA4BD' ))
```


Alternatively, a list of genes of interest can be used as input for plot_multi_genes. 
```{r fig.width=10, fig.height=7}
gene_lst <- c("IGFBP5", 'MT2A', 'TFF3', 'TFF1', 'C1QBP', 'KDM5B')

plot_multi_genes(gene_lst, binned_files_all, 'all', legend=T, colour_by_patient=T)
```


